Practical Dependently-Typed Programming for the Web

### 2013-11-11 paper/protocols.tex
Added paper revisions so far Also started the rewrite of the DB section with the new DB stuff

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
%\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\newcommand{\idata}{\textsf{iData}}
\newcommand{\itasks}{\textsf{iTasks}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

\title{Practical Dependently-Typed Programming for the Web}
\subtitle{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

\maketitle

\begin{abstract}
Dependently-typed languages allow very expressive types to be used during development, in turn facilitating easier reasoning about the operation of programs written in such languages. Stronger type specifications do however bring with them the disadvantage
that it becomes increasingly difficult to write programs that are accepted by
the type checker and additional proofs may have to be specified by a user.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in high-level languages which use dependent
types to enforce certain invariants without additional proof obligations.

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

%\keywords
%keyword1, keyword2

\section{Introduction}

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches using such exploits are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. A significant drawback, however, is that such languages
do not provide  the same static guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

%%------- EXAMPLE
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

%
\begin{Verbatim}
  $conn = new mysqli("localhost", "username", 
                     "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it. However, this would cause
problems with later steps.

%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the
\texttt{`dept'} value:

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. 
Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.

%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be
closed in order to discard the associated resources:

%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
The difficulty is 
that automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.

\idris{} \cite{brady2011idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

\subsection{Contributions}
The primary contribution of this paper is the application of 
dependent types to provide strong static guarantees
about the correctness and security of web applications, whilst minimising
additional development complexity. In particular, we present:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection and cross-site scripting.

\item Representations of CGI, Databases and sessions as
\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A message board application, demonstrating the usage of the framework.

\end{itemize}

We structure the remainder of this paper as follows. We provide a brief
overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
how this may be used to ensure adherence to resource usage protocols for CGI,
SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
type-safe form handling in Section ~\ref{form}, implemented using
\texttt{Effects}; and discuss the larger example
of a message board system making use of these components in Section
~\ref{messageboard}.

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} framework}
\label{effects}

\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
side-effects such as state, exceptions, and I/O as \emph{algebraic
effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
by an input and output \emph{state}, which permits effectful programs to track
the progress of a resource usage protocol. Effectful programs are written
in a monadic style, with \texttt{do}-notation, with their type stating which
specific effects are available.
Effectful
programs are described using the following data type,
in the simplest case:

\begin{Verbatim}
Eff : (m : Type -> Type) ->
      (es : List EFFECT) -> (a : Type) -> Type
\end{Verbatim}

\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
which describes the context in which the effectful program will be run, a
list of side effects \texttt{es} that the program is permitted to use, and the
program’s return type \texttt{a}. The name \texttt{m} for the computation context is
suggestive of a monad, but there is no requirement for it to be so.

For example, the following type carries an integer state,
throws an exception of type \texttt{String} if the state reaches 100, 
and runs in a \texttt{Maybe} context:

\begin{Verbatim}
addState : Eff Maybe [STATE Int, EXCEPTION String] ()
addState = do val <- get
              when (val == 100) (raise "State too big")
              put (val + 1)
\end{Verbatim}

\subsection{Implementing Effects}

Effects such as state and exception are described as algebraic data types,
and run by giving \emph{handlers} for specific computation contexts. 
Effects have a corresponding \emph{resource} (in the case of state, the
resource is simply the current state). Executing an effectful operation may
change the resource and return a value:

\begin{Verbatim}
Effect : Type
Effect = (in_res : Type) -> (out_res : Type) -> 
         (val : Type) -> Type
\end{Verbatim}

For example, the state effect is described as follows:

\begin{Verbatim}
data State : Effect where
     Get :      State a a a
     Put : b -> State a b ()
\end{Verbatim}

That is, \texttt{Get} returns a value of type \texttt{a} without updating
the resource type. \texttt{Put} returns nothing, but has the effect of updating
the resource. In order to make an effect usable, we implement a handler
for a computation context by making an instance of the following class:

\begin{Verbatim}
class Handler (e : Effect) (m : Type -> Type) where
     handle : res -> 
              (eff : e res res' t) -> 
              (k: res' -> t -> m a) -> m a
\end{Verbatim}

The \texttt{handle} function takes the input resource, an effect which may
update that resource and execute a side-effect, and a continuation \texttt{k}
which takes the updated resource and the return value of the effect. We use
a continuation here primarily because there is no restriction on the number of
times a handler may invoke the continuation (raising an exception, for example,
will not call the continuation). Reading and updating states in handled
for all computation contexts \texttt{m} as follows:

\begin{Verbatim}
instance Handler State m where
     handle st Get     k = k st st
     handle st (Put n) k = k n ()
\end{Verbatim}

Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
the \texttt{Get} and \texttt{Put} operations into functions in \texttt{Eff}, as
follows:

\begin{Verbatim}
STATE : Type -> EFFECT
STATE t = MkEff t State

get : Eff m [STATE x] x
get = Get

put : x -> Eff m [STATE x] ()
put val = Put val
\end{Verbatim}

A concrete effect is simply an algebraic effect type paired with its current
resource type. This, and other
technical details, are explained in full elsewhere~\cite{brady:effects}.
For the purposes of this paper, it suffices to know how to describe and
handle new algebraic effects.

\subsection{Resource Protocols as Effects}

More generally, a program might modify the set of effects available.
This might be desirable for several reasons, such as adding a new
effect, or to update an index of a dependently typed state. In this
case, we describe programs using the \texttt{EffM} data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) -> (es' : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}

\texttt{EffM} is parameterised over the context and type as before, but
separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
In fact, \texttt{Eff}
is defined in terms of \texttt{EffM}, with equal input/output effects.
We can use this to describe how effects follow resource protocols. A simple
example is a file access protocol, where a file must be opened before it
is read or written, and a file must be closed on exit:

\begin{Verbatim}
data FileIO : Effect where
     Open  : String -> (m : Mode) -> 
             FileIO () (OpenFile m) ()
     Close : FileIO (OpenFile m) () ()

     ReadLine  : FileIO (OpenFile Read)  
                        (OpenFile Read) String
     WriteLine : String -> 
                 FileIO (OpenFile Write) 
                        (OpenFile Write) ()
     EOF       : FileIO (OpenFile Read)  
                        (OpenFile Read) Bool
\end{Verbatim}

Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file changes an empty resource to
a resource containing an open file; reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions, as well as throwing an exception
if any operation fails.

The following program type checks, and therefore implicitly carries 
a proof that the file resource protocol is followed correctly:

\begin{Verbatim}
testFile : Eff IO [FILE_IO (), STDIO] () 
testFile = catch (do open "testFile" Read
                     str <- readLine
                     close
                     putStrLn str)
                 (\err => putStrLn ("Error: " ++ show err))
\end{Verbatim}

The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit.
Therefore, attempting to write to the file, or failing to open or close the
file, would cause a \emph{compile-time} error.

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effect, implicitly carry a proof that the resource
protocol has been correctly followed.

% =================================================
% =================================================

@@

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().
\begin{Verbatim}[samepage]
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a resource-dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource.

@@

\begin{figure}[h]
\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
\caption{CGI Effect}
\label{fig:cgieffect}
\end{figure}

@@

In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We may therefore define the following type synonym, which we may use in order to make use of the CGI framework: 
\begin{Verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{Verbatim}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

A simple ``Hello, world!'' program would be defined as follows:

\begin{Verbatim}
module Main
import Cgi

@@

main : IO ()
main = runCGI [initCGIState] sayHello
\end{Verbatim}
% =================================================

@@

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExecuting};

@@

\begin{figure}[h]
\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
\caption{Database Effect}
\label{fig:dbeffect}
\end{figure}

There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

data DBPointer = ValidConn Ptr
               | InvalidConn

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{Verbatim}
If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor. 
\begin{Verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{Verbatim}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

@@

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

>>>>

@@

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().

\begin{SaveVerbatim}{cgistep}

data CGIStep = Initialised   | TaskRunning 
             | TaskCompleted | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
     ICgi : CGIInfo -> InitialisedCGI s

\end{SaveVerbatim}
\useverb{cgistep}

\noindent
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a resource-dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource. The \texttt{Cgi}
effect declaration shows these resource updates in the types of each operation,
effectively specifying a state machine.

@@

\begin{SaveVerbatim}{cgieff}
{-                     { Input resource type }         { Output resource type }        { Value } -}

data Cgi : Effect where
    Init         : Cgi ()                              (InitialisedCGI Initialised)    ()
    StartRun     : Cgi (InitialisedCGI Initialised)    (InitialisedCGI TaskRunning)    ()
    FinishRun    : Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskCompleted)  ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted)  (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) (InitialisedCGI ContentWritten) ()
    OutputData   : String -> 
                   Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskRunning)    ()
    RunAction    : Env IO (CGI (InitialisedCGI TaskRunning) :: effs) -> CGIProg effs a -> 
                   Cgi (InitialisedCGI TaskRunning)    (InitialisedCGI TaskRunning)    a
\end{SaveVerbatim}

\begin{figure*}[t]
\begin{center}
\useverb{cgieff}
\end{center}
\caption{CGI Effect}
\label{fig:cgieffect}
\end{figure*}

@@

In \idris{}, types are first-class, meaning that they may be treated like other
terms in computations. We may therefore define the following type synonym,
which we may use in order to make use of the CGI framework:

\begin{SaveVerbatim}{cgiprog}

CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
  Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a

\end{SaveVerbatim}
\useverb{cgiprog}

\noindent
This is then passed, along with initial values for other effects that the user
may wish to use, to the runAction function, which invokes the RunAction
operation and executes the user-specified action.
%
A simple ``Hello, world!'' program would be defined as follows:

\begin{SaveVerbatim}{hworldcgi}

module Main
import Cgi

@@

main : IO ()
main = runCGI [initCGIState] sayHello

\end{SaveVerbatim}
\useverb{hworldcgi}
% =================================================

@@

@@

\begin{SaveVerbatim}{databaseeff}
{-                           { Input resource type }    { Output resource type }             { Value }           -}
data Sqlite : Effect where
  OpenDB            : DBName -> 
                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
  PrepareStatement  : QueryString -> 
                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
  BindInt           : ArgPos -> Int -> 
                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
                                                          (SQLitePSSuccess Bound))           (Either BindError ())
  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow)         StepResult
  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow))        StepResult
  GetColumnText     : Column -> 
                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String

\end{SaveVerbatim}

\begin{figure*}[t]
\begin{center}
\useverb{databaseeff}
\end{center}
\caption{Database Effect}
\label{fig:dbeffect}
\end{figure*}

%TODO: Fix up the labels
\begin{figure}[h]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExec};

\noindent
There are three main phases involved in the usage of the SQLite protocol:
connection to the database, preparation of a query, and execution of the query.
We define several resources to encapsulate the state at any given point during the protocol.

We first define the \texttt{SQLiteConnected} resource, which signifies that a successful connection has been made to the database. This resource contains a pointer to the database structure which is used in further computations.

\begin{SaveVerbatim}{sqliteconnected}

data SQLiteConnected : Type where
  SQLConnection : ConnectionPtr -> SQLiteConnected

\end{SaveVerbatim}  
\useverb{sqliteconnected}

\noindent
We secondly define the \texttt{SQLitePSSuccess}, \texttt{SQLitePSFail}, and \texttt{SQLiteFinishBindFail} resources to represent the process of creating and binding variables to a prepared statement.

\begin{SaveVerbatim}{sqliteconnected}

data BindStep = Binding | Bound

data SQLitePSSuccess : BindStep -> Type where
  SQLitePS       : ConnectionPtr -> 
                   StmtPtr -> SQLitePSSuccess a
  SQLiteBindFail : ConnectionPtr -> 
                   StmtPtr -> 
                   BindError -> SQLitePSSuccess a

data SQLitePSFail : Type where
  PSFail : ConnectionPtr -> SQLitePSFail

data SQLiteFinishBindFail : Type where
  SQLiteFBFail : ConnectionPtr -> StmtPtr -> 
                 SQLiteFinishBindFail

\end{SaveVerbatim}
\useverb{sqliteconnected}

\noindent
The \texttt{SQLitePSSuccess} resource indicates that a prepared statement has been correctly created by the underlying library, given a query string. The resource is parameterised by the \texttt{BindStep} type, which indicates whether the binding process has been completed. Should the creation of a prepared statement fail, the resource will be set to \texttt{SQLitePSFail}.

To avoid having to explicitly check for failures each time a variable is bound to an argument in the prepared statement, any failures are recorded by constructing the resource with the \texttt{SQLiteBindFail} constructor. Should any binding operations be called whilst in this state, the operations will not be passed to the underlying library. After the binding stage has completed, the \texttt{FinishBind} operation is called. This operation inspects the resource and either transitions to the \texttt{SQLitePSSuccess Bound} state, indicating that all bind operations were successful, or to the \texttt{SQLiteFinishBindFail} state, should a binding operation have been unsuccessful.

SQLite operates by loading database rows into a buffer, the contents of which may then be accessed through several column access functions. After all rows returned by a query have been processed, no further calls to fetch more rows may be made. Additionally, no calls to column access functions may be made whilst there is no row being currently processed.

Through the use of dependent types, we may encode these invariants statically, using the \texttt{SQLiteExecuting} resource. This is parameterised by the \texttt{ExecutionResult} data type, which indicates whether there is currently a valid row in the buffer.

\begin{SaveVerbatim}{sqliteexecuting}

data ExecutionResult = ValidRow
                     | InvalidRow

data SQLiteExecuting : ExecutionResult -> Type where
  SQLiteE : ConnectionPtr -> 
            StmtPtr -> SQLiteExecuting a

\end{SaveVerbatim}
\useverb{sqliteexecuting}

Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.

\begin{SaveVerbatim}{getcoltext}

GetColumnText : Column -> 
                Sqlite (SQLiteExecuting ValidRow) 
                       (SQLiteExecuting ValidRow)
                       String

\end{SaveVerbatim}
\useverb{getcoltext}

\noindent
In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct. The \texttt{nextRow} function 
%If a failure happens at any point during the computation, the resource is
%updated to reflect the failure: if, for example, the library failed to create a
%connection to the database, the resource value would be updated to
%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
%requests are made to the underlying SQLite library, in order to ensure safety.
%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
%\texttt{executeFail} utility functions allow for failures, once detected, to be
%handled by executing the appropriate sequence of state transition functions to
%dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
%SQL queries are evaluated in SQLite upon a call to the C library function
%\texttt{sqlite3\_step()}. In the case that a statement returns a result set,
%each subsequent call retrieves another row for processing using a column access
%function. Once all rows have been retrieved, the library returns
%\texttt{SQLITE\_DONE}, meaning that no further calls should be made without
%resetting the function. We encapsulate this requirement through the
%\texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor.

%\begin{SaveVerbatim}{stepres}

%data StepResult = Unstarted    | StepFail
%                | StepComplete | NoMoreRows

%\end{SaveVerbatim}
%\useverb{stepres}

\noindent
Each call to \texttt{nextRow}, which is a wrapper around the underlying
\texttt{sqlite3\_step()} C library function, returns a result of type
\texttt{StepResult}, which is then reflected in the resource. Calls to
\texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult}
is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore
statically guarantee that only calls that will return a valid result are
executed.

@@

\subsubsection{Example}

Programs making use of the DSL should look familiar to developers even without
a background in functional programming. To demonstrate the functionality of the
library, we return to the previous example of selecting the names and addresses
of all staff working in a given department. Due to the fact that the
\texttt{Effects} library overloads the bind operator, we may make use of
\texttt{do}-notation, facilitating the usage of an imperative style.
We define a function \texttt{textSel} of type:

\noindent
\begin{SaveVerbatim}{sqleffty}

String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))

\end{SaveVerbatim}
\useverb{sqleffty}

\noindent
This means that the program will be run in \texttt{IO}, and must
start and end with no active resources. The return type indicates that either a
list of \texttt{(String, String)} pairs, representing names and addresses in the
database, or an error will be returned.
%This example is shown in Figure
%\ref{fig:testsel}.

\begin{SaveVerbatim}{testsel}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, String)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err 
\end{SaveVerbatim}

\begin{figure}[h]
\useverb{testsel}
\caption{Example SQL program}
\label{fig:testsel}
\end{figure}

The program initially attempts to open a connection to the \texttt{people.db}
database. At this point, since the \texttt{OpenDB} operation has been invoked,
the program transitions to the \texttt{ConnectionOpened} state. The
\texttt{openDB} function returns a Boolean value indicating whether or not the
operation is successful. If not, then the \texttt{connFail} function is called
to generate an appropriate error and dispose of the resources, as shown in
~\ref{fig:sqlitestates}.
